---
title: Accessing data from the Human Cell Atlas (HCA)
teaching: 20 # Minutes of teaching in the lesson
exercises: 10 # Minutes of exercises in the lesson
---

:::::::::::::::::::::::::::::::::::::: questions

- How to obtain single-cell reference maps from the Human Cell Atlas?

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: objectives

- Learn about different resources for public single-cell RNA-seq data.
- Access data from the CELLxGENE data portal using the `cellxgenedp` package.
- Query for datasets of interest and download them into a `SingleCellExperiment` object.

::::::::::::::::::::::::::::::::::::::::::::::::


# Single Cell data sources

## HCA Project

The Human Cell Atlas (HCA) is a large project that aims to learn from and map
every cell type in the human body. The project extracts spatial and molecular
characteristics in order to understand cellular function and networks. It is an
international collaborative that charts healthy cells in the human body at all
ages. There are about 37.2 trillion cells in the human body. To read more about
the project, head over to their website at https://www.humancellatlas.org.

## CELLxGENE

CELLxGENE is a database and a suite of tools that help scientists to find,
download, explore, analyze, annotate, and publish single cell data. It includes
several analytic and visualization tools to help you to discover single cell
data patterns. To see the list of tools, browse to
https://cellxgene.cziscience.com/.

## CELLxGENE | Census

The Census provides efficient computational tooling to access, query, and
analyze all single-cell RNA data from CZ CELLxGENE Discover. Using a new access
paradigm of cell-based slicing and querying, you can interact with the data
through TileDB-SOMA, or get slices in AnnData or Seurat objects, thus
accelerating your research by significantly minimizing data harmonization at
https://chanzuckerberg.github.io/cellxgene-census/.

## The cellxgenedp package

`cellxgenedp` is a Bioconductor package that provides an R interface to the
CELLxGENE data portal. It exposes the portal's metadata as ordinary data
frames, so that datasets can be discovered with the same `dplyr` verbs used
elsewhere in this lesson, and then downloaded as `SingleCellExperiment`
objects.

The portal organizes data into a three-level hierarchy, and `cellxgenedp` gives
you one table per level:

* **collections** — a publication or project, e.g. a lung cell atlas paper
* **datasets** — the individual data objects within a collection, annotated
  with the tissues, assays, cell types, diseases and donors they contain
* **files** — the downloadable representations of a dataset (`H5AD` for
  Python/AnnData, `RDS` for Seurat)

::: callout

Earlier versions of this episode used the `CuratedAtlasQueryR` package, which
offered *cell*-level queries against a harmonized version of the same data.
That package has been deprecated and removed from Bioconductor, so this episode
now uses `cellxgenedp`. The trade-off is granularity: `cellxgenedp` queries
select whole **datasets**, and you filter to the cells of interest after
downloading. For cell-level slicing across the whole corpus, see the CELLxGENE
Census linked above.

:::

## Data Sources in R / Bioconductor

There are a few options to access single cell data with R / Bioconductor.

| Package | Target | Description |
|---------|-------------|---------|
| [cellxgenedp](https://bioconductor.org/packages/cellxgenedp) | [CellxGene](https://cellxgene.cziscience.com/) | Human and mouse SC data including HCA |
| [zellkonverter](https://bioconductor.org/packages/zellkonverter) | `H5AD` files | Reads the portal's `H5AD` files into `SingleCellExperiment` objects |

## Installation

```{r, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install(c("cellxgenedp", "zellkonverter"))
```

## Package load

```{r chunk-opts, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, message = FALSE, warning = FALSE)
library(BiocStyle)
```

Note that `SingleCellExperiment` is loaded first on purpose: it and `dplyr`
share several function names, and we want the `dplyr` versions to win.

```{r}
library(SingleCellExperiment)
library(zellkonverter)
library(dplyr)
library(cellxgenedp)
```

## Portal metadata

`db()` retrieves the metadata describing everything available in the portal.
The result is cached locally, so subsequent calls in the same session are fast.

```{r, message = FALSE}
db <- db()

db
```

The three tables are extracted with `collections()`, `datasets()` and
`files()`:

```{r}
collections(db)

datasets(db)

files(db)
```

Get a view of the first 10 columns of the dataset metadata with `glimpse()`

```{r}
datasets(db) |>
  select(1:10) |>
  glimpse()
```

## A tangent on the pipe operator

The vignette materials provided by `cellxgenedp` show the use of the
'native' R pipe (implemented after R version `4.1.0`). For those not familiar
with the pipe operator (`|>`), it allows you to chain functions by passing the
left-hand side as the first argument to the function on the right-hand side. It is used extensively in the `tidyverse` dialect of R, especially within the [`dplyr` package](https://dplyr.tidyverse.org/).

The pipe operator can be read as "and then". Thankfully, R doesn't care about whitespace, so it's common to start a new line after a pipe. Together these points enable users to "chain" complex sequences of commands into readable blocks.

In this example, we start with the built-in `mtcars` dataset and then filter to rows where `cyl` is not equal to 4, and then compute the mean `disp` value by each unique `cyl` value.

```{r}
mtcars |>
  filter(cyl != 4) |>
  summarise(avg_disp = mean(disp),
            .by = cyl)
```

This command is equivalent to the following:

```{r eval=FALSE}
summarise(filter(mtcars, cyl != 4), avg_disp = mean(disp), .by = cyl)
```

## Exploring the metadata

Let's examine the metadata to understand what information it contains.

Several of the most useful columns — tissue, assay, cell type, disease, sex,
and so on — describe a dataset with *many* values at once: a lung atlas
contains dozens of cell types, and a multi-donor study contains several
ethnicities. These are stored as list columns, one small data frame per
dataset:

```{r}
datasets(db) |>
    select(where(is.list)) |>
    names()

datasets(db) |>
    pull(tissue) |>
    head(2)
```

Rather than unpacking those by hand, `facets()` tallies how many datasets carry
each label. This is the equivalent of asking "what tissues do the experimental
data come from?":

```{r}
facets(db, "tissue") |>
    arrange(-n)
```

We can do the same for the assay types:

```{r}
facets(db, "assay") |>
    arrange(-n)
```

:::: challenge

Look through the full list of dataset metadata column names. Do any other
metadata columns jump out as interesting to you for your work?

```{r eval=FALSE}
names(datasets(db))
```

::::

## Downloading single cell data

To select datasets on the values inside those list columns, use
`facets_filter()` inside `filter()`. For the sake of demonstration, we'll focus
on datasets meeting the following criteria:

* lung tissue
* 10x 3' v3 assay
* containing CD4-positive, alpha-beta T cells

```{r}
lung_cd4 <- datasets(db) |>
    filter(
        facets_filter(tissue, "label", "lung"),
        facets_filter(assay, "label", "10x 3' v3"),
        facets_filter(cell_type, "label", "CD4-positive, alpha-beta T cell")
    )

lung_cd4 |>
    select(dataset_id, title, cell_count)
```

Out of the `r nrow(datasets(db))` datasets in the portal, `r nrow(lung_cd4)`
meet these criteria.

Each dataset is available as one or more files. We join our selection to the
file table, keep the `H5AD` representation, and pick the smallest one so that
the download finishes quickly:

```{r}
selected_file <- lung_cd4 |>
    select(dataset_id, title, cell_count) |>
    left_join(files(db) |> filter(filetype == "H5AD"),
              by = "dataset_id") |>
    arrange(filesize) |>
    slice(1)

selected_file |>
    select(title, cell_count, filesize)
```

`files_download()` fetches the file and caches it locally, so a second call
does not re-download it:

```{r, message = FALSE}
local_file <- selected_file |>
    files_download(dry.run = FALSE)

basename(local_file)
```

Finally, `readH5AD()` from `r Biocpkg("zellkonverter")` reads the file into a
`SingleCellExperiment`. Setting `use_hdf5 = TRUE` leaves the counts on disk
rather than loading them into memory, exactly as we saw in the
[Working with large data](https://carpentries-incubator.github.io/bioc-scrnaseq/large_data.html)
episode.

```{r}
sce <- readH5AD(local_file, use_hdf5 = TRUE, reader = "R")

sce
```

The cell-level annotations that the portal curates are in the `colData`, which
is where we can finally get down to individual cells — here, the CD4 cells we
selected the dataset for:

```{r}
colData(sce) |>
    as.data.frame() |>
    count(cell_type) |>
    arrange(-n) |>
    head()
```

## Save your `SingleCellExperiment`

Once you have a dataset you're happy with, you'll probably want to save it. The recommended way of saving these `SingleCellExperiment` objects is to use
`saveHDF5SummarizedExperiment` from the `HDF5Array` package.

```{r, eval=FALSE}
sce |> saveHDF5SummarizedExperiment("single_cell_counts")
```

## Exercises

:::::::::::::::::::::::::::::::::: challenge

#### Exercise 1: Basic counting + piping

Use `facets()` and `arrange()` to get the number of datasets per organism in
descending order.

:::::::::::::: solution

```{r}
facets(db, "organism") |>
    arrange(-n)
```
:::::::::::::::::::::::

:::::::::::::::::::::::::::::::::::::::::::::

:::::::::::::::::::::::::::::::::: challenge

#### Exercise 2: Counting datasets and cells

`datasets()` records the number of cells in each dataset in the `cell_count`
column. Which single collection contributes the most cells to the portal?

:::::::::::::: solution

```{r}
datasets(db) |>
    summarise(cells = sum(cell_count), .by = collection_id) |>
    arrange(-cells) |>
    slice(1) |>
    left_join(collections(db), by = "collection_id") |>
    select(name, cells)
```
:::::::::::::::::::::::

:::::::::::::::::::::::::::::::::::::::::::::

:::::::::::::::::::::::::::::::::: challenge

#### Exercise 3: Comparing metadata categories

The portal annotates both `tissue` (where the sample came from) and
`cell_type` (what the cells are). Use `facets()` to summarise each, and spot a
case where the two vocabularies describe overlapping biology.

:::::::::::::: solution

```{r}
facets(db, "tissue") |>
    arrange(-n) |>
    head()

facets(db, "cell_type") |>
    arrange(-n) |>
    head()
```

For example, blood is one of the most heavily sampled tissues, and several of
the most common cell types are the lymphocytes you would find in it.

To see the full list of metadata available for each dataset, see the
`?datasets` documentation page.

:::::::::::::::::::::::

:::::::::::::::::::::::::::::::::::::::::::::

:::::::::::::::::::::::::::::::::: challenge

#### Exercise 4: Highly specific dataset groups

Now that we are a little familiar with navigating the metadata, let's find the
datasets of 10x scRNA-seq data from lung tissue of female donors with
`COVID-19`. How many are there, and how many cells do they contain in total?

:::::::::::::: solution

```{r}
covid_lung <- datasets(db) |>
    filter(
        facets_filter(tissue, "label", "lung"),
        facets_filter(disease, "label", "COVID-19"),
        facets_filter(sex, "label", "female"),
        facets_filter(assay, "label", "10x 3' v3")
    )

nrow(covid_lung)

covid_lung |>
    summarise(datasets = n(), cells = sum(cell_count))
```

You can see we don't get very many datasets given the strict set of conditions
we used. Note that these filters select datasets that *contain* female donors
and COVID-19 samples — the downloaded object will still need subsetting to
those cells.
:::::::::::::::::::::::

:::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: keypoints

- The `cellxgenedp` package provides programmatic access to the CELLxGENE data portal, which includes data from the Human Cell Atlas.
- Portal metadata is organized as collections, datasets, and files; `facets()` summarises the multi-valued annotations attached to each dataset.
- `files_download()` retrieves a dataset's `H5AD` file, which `zellkonverter::readH5AD()` reads into a `SingleCellExperiment` object.

::::::::::::::::::::::::::::::::::::::::::::::::


## Session Info

```{r, tidy=TRUE}
sessionInfo()
```
